---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r LoadAndPreProcessData, echo=TRUE}

# In this step we will unzip the activity.zip and load activity.csv file 
unzip ("activity.zip", overwrite = TRUE);
activity_info <-read.csv("activity.csv")
summary(activity_info)

```



## What is mean total number of steps taken per day?
```{r SummarizeAndCalculateMeanMedian, echo=TRUE}
library(dplyr)

## use only those observations that have none na values for the step out
activity_info_without_na <- activity_info[!is.na(activity_info$steps),]

day_groups <- group_by(activity_info_without_na, date)
summary_by_day <- summarize(day_groups, num.types=n(), TotalSteps=sum(steps))

summary_by_day_df <- data.frame(summary_by_day)
# histogram of total steps
hist(summary_by_day_df$TotalSteps, xlab="Total Steps", main = "Histogram of Total Steps")


 mean_summary_by_day <- mean(summary_by_day_df$TotalSteps)

 median_summary_by_day <- median(summary_by_day_df$TotalSteps)

```

1. Mean Steps taken by day: `r mean_summary_by_day`
2. Median Steps taken by day: `r median_summary_by_day`


## What is the average daily activity pattern?

``` {r AverageDailyActivityPattern,echo=TRUE }

## Summarize the average data for each interval
interval_groups <- group_by(activity_info_without_na, interval)
summary_by_interval <- summarize(interval_groups, AverageSteps=mean(steps))

# create date_time value using strptime
summary_by_interval$date_time <- strptime(sprintf("%04d",summary_by_interval$interval), "%H%M")
summary_by_interval_df <- data.frame(summary_by_interval)


## Plot using ggplot2
library(ggplot2)
library(scales)

ggplot(data=summary_by_interval_df, aes(x=date_time, y=AverageSteps))+geom_line(color="blue") + xlab("Interval") + ylab("Average Steps") + ggtitle("Average Steps for each Time Interval")  + scale_x_datetime(labels = date_format("%H:%M"),breaks = "4 hour")



## pick the interval that has the max steps:
max_steps_interval <- summary_by_interval_df[summary_by_interval_df$AverageSteps == max(summary_by_interval_df$AverageSteps),1] 
```

1. Interval that has Max Steps is `r max_steps_interval`


## Imputing missing values

```{r ImputtingMissing Values, echo= TRUE}

### reporting missing values

## count of rows with na values in the original dataset
na_row_count <- nrow(activity_info[is.na(activity_info$steps),])
```
### Strategy to fix NA Values:
 * We will use the average steps for the interval across all days to fill the NA values
 * To fix an interval say interval - 110 that has an NA value, we will use the average value for this interval that has been calculated in Question 3
 * merge activity_info ( which is original data with na) to summary_by_interval_df which has the average value for each interval.

```{r ImputValues, echo=TRUE}
merged_data <- merge(activity_info,summary_by_interval_df, by="interval")

## create a new column ImputtedSteps that will have the average steps if actual steps is na
merged_data <- mutate(merged_data, ImputtedSteps = ifelse( is.na(steps) , AverageSteps ,steps))


activity_info_clean <- merged_data[,c(6,3,1)]

### match the column names to the original column names
colnames(activity_info_clean) <- colnames(activity_info)



day_groups_imputted <- group_by(activity_info_clean, date)
summary_by_day_imputted <- summarize(day_groups_imputted, num.types=n(), TotalSteps=sum(steps))

summary_by_day_imputted_df <- data.frame(summary_by_day_imputted) 

```

```{r HistogramMeanAndMedian, echo=TRUE}
# histogram of total steps of Imputted Dataset and mean and median


hist(summary_by_day_imputted_df$TotalSteps , xlab="Total Steps", main = "Histogram of Total Steps from Imputted Dataset")



 
## mean and median of imputted data
mean_imput <- mean(summary_by_day_imputted$TotalSteps)

median_imput <- median(summary_by_day_imputted$TotalSteps)

## Mean is the same for both
mean_diff <- mean(summary_by_day_imputted$TotalSteps) - mean(summary_by_day$TotalSteps)



## Median seems to be slightly higher using imputted values.

median_diff <- median(summary_by_day_imputted$TotalSteps) - median(summary_by_day$TotalSteps)

```

1. Rows with NA values in the original dataset is: `r na_row_count`
2. Mean after imputing the values: `r mean_imput`
3. Median after imputing the missing values `r median_imput`
4. There seems to be no difference in the mean between original and data that had imput values `r mean_diff`
5. Median seems to be slightly higher than mean `r median_diff`






## Are there differences in activity patterns between weekdays and weekends?

```{r ActivityPatternsByDayType,echo=TRUE}

activity_info_week <- activity_info_clean

## Add day type column to denote weekday or weekend.
activity_info_week <- mutate(activity_info_week, daytype = ifelse(strftime(strptime(date, "%Y-%m-%d"), "%u") > 5, "weekend", "weekday"))

## convert to factors.
activity_info_week$daytype <- as.factor(activity_info_week$daytype)


## create a group by using interval and daytype
interval_groups_by_daytype <- group_by(activity_info_week, interval , daytype)

average_by_interval_by_daytype <- summarize( interval_groups_by_daytype, AverageSteps=mean(steps))
average_by_interval_by_daytype_df <- data.frame(average_by_interval_by_daytype)


## create a date_time column for the timeseries plot
 average_by_interval_by_daytype_df$date_time <- strptime(sprintf("%04d",average_by_interval_by_daytype_df$interval), "%H%M")
 
ggplot(data= average_by_interval_by_daytype_df, aes(x=date_time, y=AverageSteps)) + geom_line(color="blue") + facet_grid (  daytype ~ .) + xlab("Interval") + ylab("Average Steps") + ggtitle("Average Steps for each Time Interval By Day Type (Week Day/WeekEnd")  + scale_x_datetime(labels = date_format("%H:%M"),breaks = "4 hour") +
theme(strip.text.x = element_text(size=8, angle=0),                                                                                                                                                                  strip.background = element_rect(colour="red", fill="#CCCCFF"))



```

#### Looks like the average by interval follow simiar pattern of activity. 
#### The week day activity seems to have higher averages compared to weekends.
